Rising Trends in Shelter Resident Deaths in Toronto Over the Past 17 Years

### 2024-09-24 paper/paper.qmd
Formatting erros corrected.
Some formatting erros corrected.

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
library(tidyverse)
library(knitr)
analysis_data <- read_csv("/Users/fhw0912/Desktop/Deaths-of-Shelter-Residents-in-Toronto/data/analysis_data/analysis_data.csv")
```

@@

column_description <- tibble(
  Column = c("_id", "Year", "Month", "Total decedents", "Male", "Female", "Transgender/Non-binary/Two-Spirit"),
  Description = c(
    "Unique row identifier for Open Data database.",
    "The calendar year being reported on.",
    "The month being reported on.",
    "The total number of shelter residents who died in the reported month/year.",
    "The total number of male shelter residents who died in the reported month/year.",
    "The total number of female shelter residents who died in the reported month/year.",
    "The total number of transgender, non-binary, and Two-Spirit shelter residents who died in the reported month/year."
  )
)

kable(column_description, col.names = c("Column", "Description"))
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sample
#| tbl-cap: "Sample of the cleaned data"
kable(head(analysis_data))
```

## Statistical summary of the cleaned data
@tbl-sumarry provides some basic statistics about the dataset. The project spans a total of 204 months. During this period, the highest number of deaths recorded in a single month was 19, while the lowest was 0. The average number of deaths per month is 3.848. Additionally, we can infer that the data is somewhat skewed, as the highest number of deaths in a single month is 19, while the average and median number of deaths are 3 and 3.83, respectively.
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sumarry
#| tbl-cap: "Basic statistics of the cleaned data"
kable(summary(analysis_data[, c("year_month", "total_decedents")]))
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| include: false
# 将 year_month 转换为日期格式，假设 year_month 是 YYYY-MM 格式的字符型
analysis_data$year_month_date <- as.Date(paste0(analysis_data$year_month, "-01"))

# 设置起始日期为 2007-01-01
start_date <- as.Date("2007-01-01")

# 计算每个日期与起始日期之间的时间差（以天为单位）
analysis_data$days_since_start <- as.numeric(difftime(analysis_data$year_month_date, start_date, units = "weeks"))

# 使用计算出的时间差进行回归分析
m <- lm(total_decedents ~ days_since_start, data = analysis_data)
```
I used the total number of shelter resident deaths per month as the target variable to be predicted, with time and season as predictors, and performed a linear regression analysis. Since time and season are evidently highly correlated, I decided to retain only time as the final predictor for the model. After completing the simple linear regression, I obtained a positive slope, represented by a straight line (@fig-slr), indicating that the number of shelter resident deaths in Toronto has increased over time. A comparison with the actual data revealed that the real values closely align with the predictions from the SLR model, demonstrating the reliability of my model.
```{r, results='hide'}
#| warning: false
#| message: false
#| echo: false
#| label: fig-slr
#| fig-cap: "Simple linear regression"

ggplot(analysis_data, aes(x = ym(year_month), y = total_decedents)) +
  geom_point() +  # 绘制数据点
  geom_line(aes(y = fitted(m)), col = "blue", size = 1)
  labs(title = "Total Decedents vs Year/Month",
       x = "Time",
       y = "Total Decedents")
```

####################
Figure 3 displays the total number of deaths per year, along with their respective proportions. It is important to note that I removed the data for 2007 and 2024 because they do not represent complete years. From the chart, it is evident that the number of deaths between 2020 and 2023 was significantly higher, with each of these years accounting for over 10% of the total deaths during the 15-year period. The peak occurred in 2021, reaching as high as 17.8%.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: Total Decedents per Year
#| fig-cap: "Total Decedents per Year"

# 将 year_month 转换为年份
analysis_data$year <- substr(analysis_data$year_month, 1, 4)

# 按年份汇总总死亡人数
yearly_deaths <- aggregate(total_decedents ~ year, data = analysis_data, sum)

filtered_data <- yearly_deaths[!(yearly_deaths$year %in% c(2007, 2024)), ]

# 计算每年的总死亡人数占比（百分比）
filtered_data$total_decedents_pct <- filtered_data$total_decedents / sum(filtered_data$total_decedents) * 100

# 绘制带有百分比标签的柱状图
ggplot(filtered_data, aes(x = year, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 3) +  # 添加百分比标签
  labs(title = "Total Decedents per Year", x = "Year", y = "Total Decedents") +
  theme_minimal()

@@

Additionally, Figure 4 shows the average number of deaths per season over the past few years. The proportions for spring, autumn, and summer are relatively similar, ranging between 23.6% and 23.8%. However, winter stands out with a significantly higher proportion, reaching 29%, notably surpassing the other three seasons.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: Average Decedents by Season
#| fig-cap: "Average Decedents by Season"

seasonal_data <- analysis_data %>%
  group_by(season) %>%
  summarize(mean_decedents = mean(total_decedents),
            total_decedents = sum(total_decedents))

# 计算每个季节的百分比
seasonal_data$total_decedents_pct <- seasonal_data$total_decedents / sum(seasonal_data$total_decedents) * 100

# 绘制带有百分比的柱状图
ggplot(seasonal_data, aes(x = season, y = mean_decedents, fill = season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 3) +  # 添加百分比标签
  labs(title = "Average Decedents by Season", 
       x = "Season", 
       y = "Average Decedents") +
  theme_minimal()
```

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
#| echo: false
#### Workspace setup ####
library(tidyverse)
library(knitr)
#### Read data ####
analysis_data <- read_csv("/Users/fhw0912/Desktop/Deaths-of-Shelter-Residents-in-Toronto/data/analysis_data/analysis_data.csv")
```

@@

# Build a table to display columns and thier description.
column_description <- tibble(
  Column = c("_id", "Year", "Month", "Total decedents", "Male", "Female", "Transgender/Non-binary/Two-Spirit"),
  Description = c(
    "Unique row identifier for Open Data database.",
    "The calendar year being reported on.",
    "The month being reported on.",
    "The total number of shelter residents who died in the reported month/year.",
    "The total number of male shelter residents who died in the reported month/year.",
    "The total number of female shelter residents who died in the reported month/year.",
    "The total number of transgender, non-binary, and Two-Spirit shelter residents who died in the reported month/year."
  )
)

# Draw the table.
kable(column_description, col.names = c("Column", "Description"))
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sample
#| tbl-cap: "Sample of the cleaned data"

# Display few of samples.
kable(head(analysis_data))
```

## Statistical summary of the cleaned data
@tbl-sumarry provides some basic statistics about the dataset. The project spans a total of 204 months. During this period, the highest number of deaths recorded in a single month was 19, while the lowest was 0. The average number of deaths per month is 3.848. Additionally, we can infer that the data is somewhat skewed, as the highest number of deaths in a single month is 19, while the average and median number of deaths are 3 and 3.83, respectively.
```{r}
#| warning: false
#| message: false
#| echo: false
#| label: tbl-sumarry
#| tbl-cap: "Basic statistics of the cleaned data"

# Showing the summary of cleaned data.
kable(summary(analysis_data[, c("year_month", "total_decedents")]))
```

@@

```{r}
#| warning: false
#| message: false
#| echo: false
#| include: false
# Convert year_month to a date format, assuming year_month is a character string in the format YYYY-MM
analysis_data$year_month_date <- as.Date(paste0(analysis_data$year_month, "-01"))

# Set the start date to 2007-01-01
start_date <- as.Date("2007-01-01")

# Calculate the time difference (in weeks) between each date and the start date
analysis_data$days_since_start <- as.numeric(difftime(analysis_data$year_month_date, start_date, units = "weeks"))

# Perform a regression analysis using the calculated time difference
m <- lm(total_decedents ~ days_since_start, data = analysis_data)
```
I used the total number of shelter resident deaths per month as the target variable to be predicted, with time and season as predictors, and performed a linear regression analysis. Since time and season are evidently highly correlated, I decided to retain only time as the final predictor for the model. After completing the simple linear regression, I obtained a positive slope, represented by a straight line (@fig-slr), indicating that the number of shelter resident deaths in Toronto has increased over time. A comparison with the actual data revealed that the real values closely align with the predictions from the SLR model, demonstrating the reliability of my model.
```{r, results='hide'}
#| warning: false
#| message: false
#| echo: false
#| label: fig-slr
#| fig-cap: "Simple linear regression"

# Scatter plot and fitted line
ggplot(analysis_data, aes(x = ym(year_month), y = total_decedents)) +
  geom_point() +
  geom_line(aes(y = fitted(m)), col = "blue", size = 1)
  labs(title = "Total Decedents vs Year/Month",
       x = "Time",
       y = "Total Decedents")
```

@fig-yeardeath displays the total number of shelter resident deaths per year, along with their respective proportions. It is important to note that I excluded the data for 2007 and 2024, as these are incomplete years and including them could mislead the reader. From the chart, it is clear that the number of deaths between 2020 and 2023 was significantly higher, with each of these years accounting for over 10% of the total deaths during the 15-year period. The peak occurred in 2021, where the number of deaths exceeded 125, representing 17.8% of the total.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-yeardeath
#| fig-cap: "Total Decedents per Year"

# Convert year_month to year
analysis_data$year <- substr(analysis_data$year_month, 1, 4)

# Aggregate total deaths by year
yearly_deaths <- aggregate(total_decedents ~ year, data = analysis_data, sum)

# Filter out data for the years 2007 and 2024
filtered_data <- yearly_deaths[!(yearly_deaths$year %in% c(2007, 2024)), ]

# Calculate the percentage of total deaths per year
filtered_data$total_decedents_pct <- filtered_data$total_decedents / sum(filtered_data$total_decedents) * 100

# Create a bar chart with percentage labels
ggplot(filtered_data, aes(x = year, y = total_decedents)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 3) +  # Add percentage labels
  labs(title = "Total Decedents per Year", x = "Year", y = "Total Decedents") +
  theme_minimal()

@@

Additionally, @fig-seasondeath shows the average number of deaths per season over the past few years. The proportions for spring, autumn, and summer are relatively similar, ranging between 23.6% and 23.8%. However, winter stands out with a significantly higher proportion, reaching 29%, notably surpassing the other three seasons.

```{r}
#| warning: false
#| message: false
#| echo: false
#| label: fig-seasondeath
#| fig-cap: "Average Decedents by Season"

# Grouping data by season
seasonal_data <- analysis_data %>%
  group_by(season) %>%
  summarize(mean_decedents = mean(total_decedents),
            total_decedents = sum(total_decedents))

# Calculate the percentage of total deaths for each season
seasonal_data$total_decedents_pct <- seasonal_data$total_decedents / sum(seasonal_data$total_decedents) * 100

# Create a bar chart with percentage labels
ggplot(seasonal_data, aes(x = season, y = mean_decedents, fill = season)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(total_decedents_pct, 1), "%")), 
            vjust = -0.5, size = 3) +  # Add percentage labels
  labs(title = "Average Decedents by Season", 
       x = "Season", 
       y = "Average Decedents") +
  theme_minimal()
```
